---
title: "`SLDeepAssay`: A package for maximum likelihood estimation from serial dilution assays with partial deep sequencing"
author: "Sarah C. Lotspeich and Brian D. Richardson"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installation 

Installation of the `SLDeepAssay` from GitHub requires the [`devtools`](https://www.r-project.org/nosvn/pandoc/devtools.html) package and can be done in the following way. 

```{r, eval = F}
# Install the package
devtools::install_github(repo = "sarahlotspeich/SLDeepAssay", 
                         ref = "main")
```
```{r}
# Then load it
library(SLDeepAssay)
```

The `SLDeepAssay` package contains parallel functions designed to analyze (or generate) assay data from single or multiple dilution levels. A common naming convention used throughout is to use the suffixes `_sd` and `_md` to denote functions for single- and multiple-dilution assay data, respectively. 

## Built-In Functions to Simulate Assay Data

We begin by demonstrating how to simulate assay data from a single dilution level using the `simulate_assay_sd()` function. 

There are 3 required arguments, 
  
  -   `M`: the total number of wells (a scalar),
  -   `tau`: the mean counts of infected cells per well with each DVL (a vector of with all elements $>0$)
  -   `q`: the proportion of p24-positive wells to be deep sequenced (a scalar between $0$ and $1$), 
  
and 1 optional argument, 

  -   `remove_undetected`: a logical indicator for whether DVL that are not detected in any wells should be deleted (`TRUE`) or not (`FALSE`).

The function returns a list containing two versions of the simulated data, corresponding to the quantitative viral outgrowth assay (QVOA) (`$any_DVL`) and Ultra Deep Sequencing Assay of the Outgrowth Virus (UDSA) (`$DVL_specific`) measures discussed in the manuscript.  
  
For demonstration, we simulate a single-dilution assay of `M = 12` total wells (each with `u = 1` million cells in it), with `n = 6` underlying DVL, and assume that `q = 0.5` (i.e., 50\%) of p24-positive wells will undergo deep sequencing. Lastly, if we want to simulate data where the overall IUPM = 0.5, we set `tau = rep(0.5 / 6, 6)` (or any other way to distribute a total mass of 0.5 across the 6 DVL). 

We now demonstrate a few ways to simulate, depending on our choices for the optional arguments. 

```{r}
assay = simulate_assay_sd(M = 12, 
                          tau = rep(0.5 / 6, 6), 
                          q = 0.5, 
                          remove_undetected = TRUE)
assay
```

A few things to notice here... 

  -   `assay$any_DVL` is a vector containing indicators for each well, where `1` indicates that the well tested positive for at least one DVL in the QVOA. 
  -   `assay$DVL_specific` only has 2 rows when we set `n = 6`. Where did they go?? Since `remove_undetected = TRUE` here, the other 4 rows were deleted because they must not have been detected in either of the 2 p24-positive and deep sequenced wells. (We will confirm this in a moment.)
  -   `assay$DVL_specific` has columns of `NA` values at the end. This isn't an error! The wells represented by columns 10--12 are the $M_P - m$ p24-positive but not deep sequenced wells. They are correctly captured as being infected with *something* in `assay$any_DVL`, though.  

### The `remove_undetected` option

Now what if I leave the `seed` the same, but instead set `remove_undetected = FALSE`? 

```{r}
assay = simulate_assay_sd(M = 12,
                          tau = rep(0.5 / 6, 6), 
                          q = 0.5, 
                          remove_undetected = FALSE)
assay
```

Notice that `assay$any_DVL` is the same as before, which makes sense since it's the same random seed. However, `assay$DVL_specific` now has all `n = 6` rows, even though rows 1--2 and 4--5 contain all 0s. 

### Multiple dilution levels

Now if we instead want to simulate assay data from multiple dilution levels, we would use the `simulate_assay_md()` function. There are still 4 required arguments, but now they are 
  
  -   `M`: the total number of wells (a scalar),
  -   `tau`: 
  -   `q`: the proportion of p24-positive wells to be deep sequenced (a scalar between $0$ and $1$), and 
  -   `dilutions`:  
  
The 2 optional arguments are the same as with the `simulate_assay_sd()` function. 

## Built-In Functions for Estimation 

Now that we have our assay data, we turn to primary purpose of the `SLDeepAssay` package: to efficiently estimate the rate of infection from dilution assays with deep sequencing data using maximum likelihood estimation (MLE). We can obtain point estimates, standard errors, and confidence intervals all with a single call to the `fit_SLDeepAssay_sd()` function. There are actually two ways to supply the assay data for estimation. 

### Supplying full assay data 

The first argument `assay` expects a matrix like the one `simulate_assay_sd()` returned in `assay$DVL_specific` above. If this is supplied, then the only other *required* argument is `u`, which is the dilution level in millions of cells per well. Then, you can skip parameters `M`, `MP`, `m`, and `Y` (they will be discussed in the next section), and all that is left are optional arguments: 

  -   `corrected`: a logical indicator for whether the bias-corrected MLE should be returned (`TRUE`) or not (`FALSE`),
  -   `maxit`: the maximum number of iterations allowed to find the MLE,
  -   `lb`: the lower bound for the parameters, and
  -   `ub`: the upper bound for the parameters. 
  
The last three optional parameters are all passed to the `optim()` function, which is used to find the MLE inside of `simulate_assay_sd()`. Recall that data were simulated for wells with 1 million cells per well, so `u = 1` and the MLE can be interpreted as the infectious units per million (IUPM).

We now estimate the IUPM based on the last `assay` generated above, leaving all optional arguments at their defaults.

```{r}
res = fit_SLDeepAssay_sd(assay = assay$DVL_specific, 
                         u = 1)
res
```

Finally, we interpret the output as follows. 

  -   `res$mle` is the uncorrected MLE. 
  -   `res$se` is the estimated standard error (assumed to be the same for both the bias corrected and uncorrected MLE). 
  -   `res$ci` is the 95\% confidence interval (CI) for the uncorrected MLE. 
  -   `res$mle_bc` is the bias-corrected MLE. 
  -   `res$ci_bc` is the 95\% CI for the bias-corrected MLE. 

### Supplying summarized assay data
